---
title: "Capstone Decision Tree"
output: pdf_document
date: "2025-04-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Decision Tree with Original Variables 
```{r, message=FALSE, warning=FALSE}
library(readxl)
library(dplyr)
library(caret)
library(rpart)
library(rpart.plot)
df_raw <- read_excel("capstonedata.xlsx")

# Drop unnecessary columns and ensure all predictors are numeric
df <- df_raw %>%
  select(-`...1`, -X, -Country, -Year) %>%
  na.omit()

# Convert all columns except Life.Ladder to numeric
df <- df %>%
  mutate(across(-`Life.Ladder`, as.numeric))

# Create binary outcome from Life.Ladder
median_ll <- median(df$`Life.Ladder`, na.rm = TRUE)
df <- df %>%
  mutate(LifeLadderBinary = ifelse(`Life.Ladder` >= median_ll, 1, 0),
         LifeLadderBinary = as.factor(LifeLadderBinary))

# Train/test split
set.seed(123)
train_index <- createDataPartition(df$LifeLadderBinary, p = 0.7, list = FALSE)
train_data <- df[train_index, ]
test_data <- df[-train_index, ]
```
# Decision Tree with Original Variables
```{r}
# Drop target variables from predictors
predictors_orig <- train_data %>% select(-`Life.Ladder`, -LifeLadderBinary)
response_orig <- train_data$LifeLadderBinary

# Fit decision tree
tree_orig <- rpart(LifeLadderBinary ~ ., data = cbind(predictors_orig, LifeLadderBinary = response_orig), method = "class")
rpart.plot(tree_orig)

# Predict and evaluate
pred_orig <- predict(tree_orig, test_data, type = "class")
conf_orig <- confusionMatrix(pred_orig, test_data$LifeLadderBinary, positive = "1")
print(conf_orig)
```
This original variables based tree model prioritizes variables such as 
Log.GDP.per.capita, Freedom.to.make.life.choices, GDP.per.capita..current.US.., Social.support, and Healthy.life.expectancy.at.birth in determining whether a 
country falls above or below the median Life Ladder score. The first and most 
influential split occurs at Log.GDP.per.capita < 9.8, indicating that economic
prosperity is a key driver of life satisfaction. Subsequent splits reflect access
to healthcare, governance (e.g., Voice.and.Accountability..Percentile.Rank), and
tax burden, all of which align with common contributors to well-being.

# Decision Tree with Principal Components
```{r}
# Extract predictors and convert to numeric
df_pca_input <- df %>% select(-`Life.Ladder`, -LifeLadderBinary)
df_pca_input <- df_pca_input %>% mutate(across(everything(), as.numeric))

# Remove near-zero variance columns
nzv <- nearZeroVar(df_pca_input)
if (length(nzv) > 0) {
  df_pca_input <- df_pca_input[, -nzv]
}

# Remove rows with any NA or Inf
valid_rows <- apply(df_pca_input, 1, function(row) all(is.finite(row) & !is.na(row)))
df_pca_input_clean <- df_pca_input[valid_rows, ]
target_clean <- df$LifeLadderBinary[valid_rows]

# Run PCA
df_pca_scaled <- scale(df_pca_input_clean)
pca <- prcomp(df_pca_scaled, center = TRUE, scale. = TRUE)

# Keep top 10 PCs and align with clean target
pc_df <- as.data.frame(pca$x[, 1:10])
pc_df$LifeLadderBinary <- target_clean

# Train/Test split for PCA dataset
set.seed(123)
train_index_pc <- createDataPartition(pc_df$LifeLadderBinary, p = 0.7, list = FALSE)
train_pc <- pc_df[train_index_pc, ]
test_pc <- pc_df[-train_index_pc, ]

# Train decision tree
tree_pc <- rpart(LifeLadderBinary ~ ., data = train_pc, method = "class")
rpart.plot(tree_pc)

# Predict and evaluate
pred_pc <- predict(tree_pc, test_pc, type = "class")
conf_pc <- confusionMatrix(pred_pc, test_pc$LifeLadderBinary, positive = "1")
print(conf_pc)
```
The PCA-based decision tree is less interpretable due to its reliance on principal components, which are linear combinations of the original variables. However, 
the first split on PC1 still captures most of the variance in the data, and the
tree structure remains simple and effective. While individual splits can’t be 
directly tied to specific variables, the model still performed well, suggesting 
that the principal components retained meaningful patterns for classification.

# Comparison
```{r}
metrics <- data.frame(
  Model = c("Original Variables", "Principal Components"),
  Accuracy = c(0.8347, 0.8667),
  Precision = c(0.8727, 0.8636),
  Recall = c(0.7869, 0.9500)
)

print(metrics)
```
Overall, the original-variable model offers better interpretability, showing clear relationships between social and economic indicators and life satisfaction.
The PCA model, while less transparent, captures similar structure in a more 
condensed form and demonstrates slightly stronger classification performance.

The model trained on the original variables achieved an accuracy of 83.5%, with
a precision of 87.3% and a recall of 78.7%. In contrast, the model trained on 
the first ten principal components from PCA slightly outperformed in terms of 
accuracy and recall, reaching 86.7% accuracy and an impressive 95.0% recall, 
although it had a slightly lower precision of 86.4%. 