---
output:
  bookdown::pdf_document2:
    toc: false
fontsize: 12pt
geometry: margin=1in
linestretch: 2
---

\begin{titlepage}
\centering
\vspace*{4cm}

{\Huge \textbf{Models Evaluations Report} \par}
\vspace{2cm}

{\Large Anh Ly, Hayeon Chung, Aidan Loughran, and Tran Nguyen \par}
\vspace{1cm}

{\large March 24, 2025 \par}

\vfill

{\large STAT 4996 Capstone \par}
{\large Jolly Ranchers \par}

\end{titlepage}

\newpage

\tableofcontents

\newpage


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(ggplot2)
library(factoextra) 
library(dplyr)
library(caret)
library(psych)
library(pls)
library(readxl)
library(writexl)
```

# Principal Component Analysis 

```{r load process data, echo=FALSE, results='hide'}
df <- read.csv("/Users/anhly/Desktop/STAT 4996/Visualization/cleandf.csv")

# Removing non-feature columns 
df <- df %>% select(-Country, -Year)

# Replacing ".." with NA
df[df == ".."] <- NA

# Converting all columns to numeric 
df_numeric <- df %>%
  mutate(across(everything(), ~as.numeric(.)))

# Replacing NAs in numeric columns with median values
df_filled <- df_numeric %>%
  mutate(across(everything(), ~ifelse(is.na(.), median(., na.rm = TRUE), .)))

# Confirming there are no NAs left
sum(is.na(df_filled))  # output should be 0
```


```{r Define Target Variable and Split Data, echo = FALSE, results='hide'}
# Use "Life.Ladder" as the target variable
df_filled$target_variable <- df_filled$Life.Ladder
df_filled <- df_filled %>% select(-Life.Ladder)  # Removing original column to avoid duplication

# Train-test split
set.seed(123)
splitIndex <- createDataPartition(df_filled$target_variable, p = 0.7, list = FALSE)
train_data <- df_filled[splitIndex, ]
test_data <- df_filled[-splitIndex, ]
```


```{r PCA on Training Data Only, echo=FALSE, results='hide'}
# Remove target variable before PCA
train_x <- train_data %>% select(-target_variable)
test_x <- test_data %>% select(-target_variable)

# Standardize and perform PCA
pca_result <- prcomp(train_x, center = TRUE, scale. = TRUE)
summary(pca_result)

# Transform train and test sets
train_pcs <- as.data.frame(predict(pca_result, newdata = train_x))
test_pcs <- as.data.frame(predict(pca_result, newdata = test_x))

# Add target variable back
train_pcs$target_variable <- train_data$target_variable
test_pcs$target_variable <- test_data$target_variable
```

```{r fig-PCA-scree-plot, echo=FALSE, out.width='80%', fig.cap="Scree Plot for PCA 1 and PCA 2"}
fviz_eig(pca_result)
```

## PCA on Training Data 
Figure \@ref(fig:fig-PCA-scree-plot) provides a clear visual representation of how much variance each 
principal component explains. As shown, the first principal component (PC1) stands 
out by explaining approximately 26.6% of the total variance which is more than double 
that of any other component. The second component (PC2) accounts for about 10.4%, 
and together, PC1 and PC2 explain just over 37% of the variance in the dataset. 
The third (PC3) and fourth (PC4) components contribute around 7.5% and 4.8%, 
respectively, with the explained variance continuing to decline gradually thereafter.

Figure \@ref(fig:fig-PCA-scree-plot) shows a sharp “elbow” after the second component, indicating that PC1 
and PC2 capture the most meaningful structure in the data. This elbow suggests 
that a two-component solution may be effective for visualizing or summarizing 
the data in reduced dimensions.

## PCA Interpretation 
```{r PCA interpretation}
# Variable loadings
pca_loadings <- pca_result$rotation
sort(pca_loadings[, "PC1"], decreasing = TRUE)[1:10]
sort(pca_loadings[, "PC2"], decreasing = TRUE)[1:10]   
```

The first principal component appears to capture a comprehensive measure
of development and well-being, strongly driven by indicators across multiple 
domains. Within the health category, life expectancy at birth and healthy life 
expectancy are among the top contributors, highlighting the central role of 
population health outcomes. From the infrastructure development category, access 
to internet, basic sanitation, electricity, and safe drinking water also load 
heavily on PC1, indicating the importance of basic service access. Additionally,
GDP per capita and log GDP per capita from the economic indicators category, 
along with the subjective well-being measure Life Ladder, and social support
from the social indicators category, further support the interpretation of PC1 
as a representative for overall socioeconomic development and quality of life. Countries with higher PC1 scores tend to enjoy better public health, infrastructure, 
economic conditions, and general life satisfaction.

In contrast, the second principal component (PC2) emphasizes aspects related 
more to governance, representation, and institutional inclusion. Its top 
contributors include Voice and Accountability (Percentile Rank) and Proportion
of seats held by women in national parliaments, both of which fall under the 
social indicators category and reflect levels of civic engagement and gender 
equity in political participation. While several indicators from the 
infrastructure and health categories—such as access to water, internet, and 
healthy life expectancy—also contribute moderately to PC2, their influence is 
less pronounced than in PC1. The presence of the unemployment rate from the 
economic indicators category further suggests that PC2 may reflect institutional
structures that impact equity and opportunity.

## Linear Regression Using Principal Components 

```{r, echo=FALSE}
# Use the first 10 principal components to predict target_variable
pcr_model <- lm(target_variable ~ ., data = train_pcs %>% select(PC1:PC10, target_variable))
summary(pcr_model)
```

The summary output reveals that the PCR model using the first 10 principal components as predictors performs quite well in modeling the Life.Ladder outcome variable. The Multiple R-squared is 0.7895, and the Adjusted R-squared is 0.784, which indicates that nearly 79% of the variance in Life.Ladder is explained by the selected principal components. This suggests a strong fit and demonstrates that the reduced dimensional representation of the original dataset captures the key patterns related to well-being and life satisfaction.

Looking at the coefficients, we see that PC1 has the most substantial positive impact on the target variable, with a large coefficient estimate of 0.2325 and a highly significant p-value (< 2e-16). This suggests it captures critical information related to socioeconomic well-being. Interestingly, PC3 has a large negative coefficient (-0.2458), also highly significant, indicating that the variation captured by PC3 is inversely related to Life.Ladder. Other components like PC2, PC4, PC6, PC7, PC8, PC9, and PC10 are also statistically significant (p < 0.01), implying that each component adds meaningful explanatory power to the model. The residual standard error of 0.5102 suggests a reasonably tight fit of predicted values around the actual data.

## Model Evaluation on Test Set

```{r, echo=FALSE}
# Predict on test data
predictions <- predict(pcr_model, newdata = test_pcs)

# Calculate RMSE and R-squared
rmse <- sqrt(mean((predictions - test_pcs$target_variable)^2))
rsq <- 1 - sum((predictions - test_pcs$target_variable)^2) / sum((mean(train_pcs$target_variable) - test_pcs$target_variable)^2)

cat("Test RMSE:", rmse, "\n")
cat("Test R-squared:", rsq)
```
This section provides insight into how well the PCR model generalizes to unseen data. 
The RMSE quantifies the average magnitude of prediction errors. The value of 0.511 
means that, on average, the predictions made by our PCR model differ from the actual
Life.Ladder scores in the test set by about 0.51 units on a scale that ranges from
0 to 10. Since life satisfaction scores are often reported with one decimal place, this RMSE value suggests the model's predictions are reasonably accurate. 

## Plot Predictions vs Actual

```{r fig-scatterplot-pred-act, echo=FALSE}
ggplot(data.frame(Actual = test_pcs$target_variable, Predicted = predictions),
       aes(x = Actual, y = Predicted)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = 'red', linetype = 'dashed') +
  theme_minimal() +
  labs(title = "Predicted vs Actual", x = "Actual", y = "Predicted")
```

Figure \@ref(fig:fig-scatterplot-pred-act) offers a strong visual confirmation of the PCR model's performance. 
Each point represents an observation in the test set, with its actual Life.Ladder score on the x-axis and the model's predicted score on the y-axis. The points are 
clustered around the red dashed line indicating a strong alignment between predicted
and true values. Where there is some dispersion at the lower end, the overall pattern
suggests that the model is both accurate and consistent in its predictions. 

\newpage

# Composite Score 

```{r, echo=FALSE}
file <- "/Users/anhly/Desktop/STAT 4996/Analysis/data_with_comp_scores.xlsx"

scoresdata <- read_excel(file)
scoresdata[scoresdata == "NA"] <- NA

scoresdata$Health.Composite.Score <- as.numeric(as.character(scoresdata$Health.Composite.Score))
scoresdata$Education.Composite.Score <- as.numeric(as.character(scoresdata$Education.Composite.Score))
scoresdata$Social.Composite.Score <- as.numeric(as.character(scoresdata$Social.Composite.Score))
scoresdata$Economic.Composite.Score <- as.numeric(as.character(scoresdata$Economic.Composite.Score))
scoresdata$Infrastructure.Composite.Score <- as.numeric(as.character(scoresdata$Infrastructure.Composite.Score))
scoresdata$Life.Ladder <- as.numeric(as.character(scoresdata$Life.Ladder))
scores_clean <- na.omit(scoresdata)

mod1 <- lm(Life.Ladder ~ Health.Composite.Score + Education.Composite.Score + Social.Composite.Score + Economic.Composite.Score + Infrastructure.Composite.Score, data = scores_clean)

#Show regression summary
summary(mod1)
```

## Model Summary

-Adjusted R-squared: 0.7178 (indicating that about 71.78% of the variance in the Life Ladder score is explained by the composite scores)

-Residual Standard Error: 0.4208

-F-statistic: 52.38 (p-value \< 2.2e-16), indicating that the model as a whole is statistically significant.

## Interpretation

-The Infrastructure Composite Score has the highest positive impact on the Life Ladder score, followed by the Economic Composite Score.

-Education and Social Composite Scores also positively contribute but with lower statistical significance.

-Health Composite Score does not significantly impact the Life Ladder score.

# Principal Component Regression 

```{r pcr, echo=FALSE}
#copy code over (the setup)
```

## PCR Compositie Score 1
```{r}
#print out pcr comp score 1 and write blurb (interpretation)
```

## Linear Regression using PCR 
```{r}
#test on untrained data - interpretation written underneath 
```

